---
title: "Exploratory Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

Load packages and import datasets.
```{r, message = FALSE}
library(ggridges)
library(tidyverse)
library(plotly)

combine_df = read_csv("data/combine.csv") 

combine_tc_binary = read_csv("data/combine.csv") |> 
  mutate(
    total_cholesterol = case_when(
      total_cholesterol < 200 ~ "desirable",
      total_cholesterol >= 200 ~ "above desirable",
      TRUE ~ as.character(total_cholesterol)
    )
  )
```

## Labeling the bar chart representing **Total Cholesterol**

```{r, message = FALSE}
TC_distri = ggplot(combine_df, aes(x = total_cholesterol)) +
  geom_histogram(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(
    x = "Total Cholesterol",
    y = "Frequency",
    title = "Distribution of Total Cholesterol"
  ) +
  theme_minimal()

ggplotly(TC_distri)
```

The presented data illustrates the distribution of `Total Cholesterol`, revealing a notable *right-skewed graph*. This skewness is indicative of a pronounced prevalence of participants whose cholesterol levels are concentrated within the range of 130 to 200 mg/dL. 

Specifically, a substantial number of individuals exhibit cholesterol values within this interval, emphasizing the central tendency of the data around these values. The rightward skewness further implies that there are fewer participants with cholesterol levels exceeding 200 mg/dL, suggesting an asymmetry in the upper tail of the distribution. 

## Labeling the graph for **Sleep Hours** & Total Cholesterol

```{r, message = FALSE, warning = FALSE}
sleep_hour_distri = ggplot(combine_df, aes(x = sleep_hour)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Sleep Hours",
       x = "Sleep Hours",
       y = "Frequency") +
  theme_minimal()

ggplotly(sleep_hour_distri)
```

```{r, message = FALSE, warning = FALSE}
TC_sleep = 
  plot_ly(combine_tc_binary, x = ~total_cholesterol, y = ~sleep_hour, color = ~total_cholesterol, type = "box", colors = "viridis") |> 
  layout(title = "Total Cholesterol & Sleep Hours",
         xaxis = list(title = "Total Cholesterol"),
         yaxis = list(title = "Sleep Hours"))

TC_sleep
```

The distribution of `Sleep Hours` depicted above exhibits characteristics consistent with a *normal distribution*, showcasing a bell-shaped curve. This normal distribution suggests that the majority of individuals in the sample population tend to obtain approximately 7 to 8 hours of sleep per night. The peak of the curve is centered around this range, indicating a prevalent and typical sleep duration among the surveyed participants. 

The presented *boxplot* visually represents the distribution of `Sleep Hours` in conjunction with `Total Cholesterol` levels*, revealing a remarkable similarity in distribution patterns between the group categorized as "desirable" and those classified as "above desirable". This observation suggests that individuals within both categories exhibit comparable trends in sleep duration.

*`Total Cholesterol` levels: total blood cholesterol ≥ 200 mg/dL is defined as above desirable (CDC, 2023).

## Labeling the bar chart representing **Physical Actively** & Total Cholesterol

```{r, message = FALSE, warning = FALSE}
combine_tc_binary = 
  combine_tc_binary |> 
  drop_na(physical_activity)

TC_activity = ggplot(combine_tc_binary, aes(x = physical_activity, fill = total_cholesterol)) +
  geom_bar(position = "stack") +
  scale_fill_brewer(palette = "lightgray") +
  labs(
    x = "Physical Activity",
    y = "Frequency",
    fill = "Total Cholesterol",
    title = "Relationship between Physical Activity and Total Cholesterol"
  ) +
  theme_minimal()

ggplotly(TC_activity)
```

The above bar chart illustrates the association between `Physical Activity` and `Total Cholesterol`. It reveals that the majority of individuals engage in vigorous physical activity. Surprisingly, the prevalence of individuals with Total Cholesterol levels above the desirable range remains consistent across three distinct levels of Physical Activity. 