---
title: "Project Report"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
library(tidyverse)
library(haven)
library(sf)
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  fig.align = 'center')


theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

## **Motivation**
Our project aims to explore the intricate relationship between health affecting behaviors and the health outcome of High cholesterol levels. With the growing concern about cardiovascular diseases and their association with lifestyle choices, understanding these relationships is crucial. We intend to use comprehensive datasets to analyze and model these associations, providing insights that could lead to better health outcomes.

## **Related Work**
Our inspiration stems from an introductory article by the American Heart Association (2020) titled "What is Cholesterol?" This article shed light on the profound impact of cholesterol on the human body. Recognizing that daily eating habits and behaviors can contribute to elevated cholesterol levels and pose a risk to cardiovascular health, we aim to develop a website. This platform aims to empower individuals by providing insights into their cholesterol levels, educating them about the associated risks, and encouraging proactive steps to maintain or enhance their overall health.

## **Initial Questions**
The purposes of this project is to answer the following questions:

- What kind and extent of the correlations between our targeted health outcome - high cholesterol and health related behaviors at national level?
- What patterns do their correlation exhibit in the graphs?
- By using the individual dataset, which models are effective in capturing the association between high cholesterol and a set of health-related behaviors identified in national dataset that exhibit a strong correlation?
- How can we create a user-friendly tool enabling individuals to evaluate their personal risk of high cholesterol using demographic features and lifestyle choices?

## **Data Sources and Cleaning**
### *Part 1: [PLACES: Local Data for Better Health](https://data.cdc.gov/500-Cities-Places/PLACES-Local-Data-for-Better-Health-County-Data-20/swc5-untb/about_data)*
We acquire model-based county estimates of the entire United States during the time period from 2020 to 2021. The data were provided by the CDC, Division of Population Health, Epidemiology and Surveillance Branch.The estimates encompass 36 measures, spanning health outcomes, preventive services use, chronic disease-related health risk behaviors, disabilities, and health status. We are using several health-affecting behavior variables and the prevalence of High Cholesterol to evaluate and sorting out the most correlated risk factors to the outcome at the national level.

### *Part 2: [NHANCES](https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?Cycle=2017-2020)*
In order to get individual data measurements to analyze the association of risk factors and `high Cholesterol`, we use the several sub-datasets from `NHANCES` to gather all the variables we are interested from the result of part 1. This resource conducted by the CDC for the time period of 2017 to March 2020 Pre-pandemic. It combines interviews and physical examinations to collect a wide range of health-related information. Initially, all the variables were non-descriptive, we assigned descriptive names for them. While importing the data measurements, the types of all the variables were numberical. Except continuous variables, we categorizd variables `gender`, `race`, `marital_status`, `education_level`, `poverty_level`, and `alcohol_use_cat`. The numbers from drinking habit dataset are not order in a logic way, therefore we reordered variables as `alcohol_use_order`. The dataset with information of physical activity used the skip methods 
The comprehensive dataset employed for our analysis includes detail information for each participant, including key variables such as:

  - `id`: Participants' unique identification numbers
  - `total_cholesterol`: Serum total cholesterol level, in mg/dL
  - `age`: participants' Age at screening , in year
  - `gender`: 2 types: Male or Female
  - `race`: 5 types: Hispanic, Non-Hispanic White, Non-Hispanic Black, Non-Hispanic Asian, Other Race
  - `marital_status`: For the adults 20+ years old, 3 typess: Married/Living with Partner, Widowed/Divorced/Separated, Never married
  - `education_level`: For the adults 20+ years old, 4 levels: Less than 9th grade, 9-11th grade, High school graduate/GED or equivalent, Some college or AA degree, College graduate or above
  - `bmi`: Body Mass Index, in kg/m**2
  - `poverty_level`: Family monthly poverty level category, 3 levels: Below 130% of Poverty Guidelines, Between 130% and 185% of Poverty Guidelines, Above 185% of Poverty Guidelines
  - `physical_activity`: Types of weekly physical activity, 3 levels: Vigorous activity, Moderate activity, Light/Unknown activity
  - `sleep_hour`: The sleeping habits in Weekdays or workdays, in hour
  - `alcohol_use_order`: Ordinal variables, scale from 1 - 11 meaning from never to everyday consuming the alcohol in the past twelve months
  - `alcohol_use_cat`: Category of drinking patterns, 3 levels: Light Drinker, Moderate Drinker, Heavy Drinker
  - `cotinine`: Serum Cotinine concentration, in ng/mL

```{r}
tchol_df = 
  data <- read_xpt("data/xpt/P_TCHOL.XPT") |> 
  janitor::clean_names() |>
  select(id = seqn, total_cholesterol = lbxtc) |> 
  drop_na()

demo1_df = 
  data <- read_xpt("data/xpt/P_DEMO.XPT") |> 
  janitor::clean_names() |> 
  select(id = seqn, age = ridageyr, gender = riagendr, race = ridreth3, 
         marital_status = dmdmartz, education_level_20 = dmdeduc2) |> #only include the variables of id, age, gender, race, marital status, and education level.
  filter(age >= 18) |> # Only include the participants with age equal and greater than 18.
  mutate(
    gender = case_when(
      gender == 1 ~ "Male",
      gender == 2 ~ "Female",
      TRUE ~ as.character(gender)), #categorize variable gender.
    race = case_when(
      race %in% c(1, 2) ~ "Hispanic",
      race == 3 ~ "Non-Hispanic White",
      race == 4 ~ "Non-Hispanic Black",
      race == 6 ~ "Non-Hispanic Asian",
      race == 7 ~ "Other Race",
      TRUE ~ as.character(race)), #categorize variable race.
    marital_status = case_when(
      marital_status %in% c(77, 99, ".") ~ NA_character_,
      marital_status == 1 ~ "Married/Living with Partner",
      marital_status == 2 ~ "Widowed/Divorced/Separated",
      marital_status == 3 ~ "Never married",
      TRUE ~ as.character(marital_status)), #categorize variable marital status.
    education_level_20 = case_when(
      education_level_20 %in% c(7, 9, ".") ~ NA_character_,
      education_level_20 == 1 ~ "Less than 9th grade",
      education_level_20 == 2 ~ "9-11th grade",
      education_level_20 == 3 ~ "High school graduate/GED or equivalent",
      education_level_20 == 4 ~ "Some college or AA degree",
      education_level_20 == 5 ~ "College graduate or above",
      TRUE ~ as.character(education_level_20)) #categorize education level status.
   )

bmi_df = 
  data <- read_xpt("data/xpt/P_BMX.XPT") |> 
  janitor::clean_names() |>
  select(id = seqn, bmi = bmxbmi) |> 
  drop_na()

income_df = 
  read_xpt("data/xpt/P_INQ.XPT") |> 
  janitor::clean_names() |>
  select(id = seqn, poverty_level = indfmmpi) |> 
  drop_na() |> 
  mutate(
    poverty_level = case_when(
      poverty_level < 1.3 ~ "Below 130% of Poverty Guidelines",
      between(poverty_level, 1.3, 1.85) ~ "Between 130% and 185% of Poverty Guidelines",
      poverty_level > 1.85 ~ "Above 185% of Poverty Guidelines",
      TRUE ~ as.character(poverty_level))
        )

demo_df <- demo1_df |> 
  left_join(income_df, by = "id") |> 
  left_join(bmi_df, by = "id") |> 
  drop_na(age, gender, race, bmi) 

activity_df = 
  data <- read_xpt("data/xpt/P_PAQ.XPT") |> 
  janitor::clean_names() |>
  select(id = seqn, paq605, paq650, paq620, paq635, paq665) |> 
  drop_na() |> 
  mutate(
    physical_activity = case_when(
      paq605 == 1 | paq650 == 1 ~ "Vigorous activity",
      paq620 == 1 | paq635 == 1 | paq665 == 1 ~ "Moderate activity",
      paq665 == 2 ~ "Light/Unknown activity",
      TRUE ~ NA_character_ )
        ) |> 
  select(id, physical_activity)

sleep_df = 
  data <- read_xpt("data/xpt/P_SLQ.XPT") |> 
  janitor::clean_names() |>
  select(id = seqn, sleep_hour = sld012) |> 
  drop_na()

alcohol_df = 
  data <- read_xpt("data/xpt/P_ALQ.XPT") |> 
  janitor::clean_names() |>
  mutate(
    alcohol_use_cat = case_when(
      alq121 %in% c(0, 10, 9, 8) ~ "Light Drinker",
      alq121 %in% c(7, 6, 5, 4) ~ "Moderate Drinker",
      alq121 %in% c(3, 2, 1) ~ "Heavy Drinker",
      alq121 %in% c(77, 99, ".") ~ NA_character_),
     alcohol_use_order = case_when(
      alq121 == 0 ~ 1,
      alq121 == 10 ~ 2,
      alq121 == 9 ~ 3,
      alq121 == 8 ~ 4,
      alq121 == 7 ~ 5,
      alq121 == 6 ~ 6,
      alq121 == 5 ~ 7,
      alq121 == 4 ~ 8,
      alq121 == 3 ~ 9,
      alq121 == 2 ~ 10,
      alq121 == 1 ~ 11)
        ) |> 
   select(id = seqn, alcohol_use_cat, alcohol_use_order) |> 
   drop_na()

smoking_df = 
  data <- read_xpt("data/xpt/P_COT.XPT") |> 
  janitor::clean_names() |>
  select(id = seqn, cotinine = lbxcot) |> 
  drop_na()

combine_df = 
  tchol_df |> 
  left_join(demo_df, by = "id") |>
  left_join(activity_df, by = "id") |>
  left_join(sleep_df, by = "id") |>
  left_join(alcohol_df, by = "id") |>
  left_join(smoking_df, by = "id") |> 
  drop_na(age, gender, race)

write.csv(combine_df, file = "data/combine.csv")

combine_df %>% 
  knitr::kable(digits = 3) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), font_size = 12) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```
<br>
Above is the final dataset `combine_df`, with 8161 observations and 14 variables will be used in following steps of exploratory and statistical analysis. 


